Survival in chessland

### 2019-02-17 trunk/chess/papers/survival.tex
Paper editing. Code to make permutation tables. Some other miscellaney, apparently

If you are forced to play chess to the death, you are in trouble,
because most people are not good at chess (for example, the author)
and yet want to live.\footnote{It is easy for two players to
  collaborate to produce a draw, especially by simply agreeing to a
  draw at the outset of the game (if allowed). Some tournament formats
  forbid the players from agreeing to a draw verbally before a certain
  point (e.g. 30 moves). There are always other routes to a draw, for
  example by stalemate or repeating the same position three times.
  Collaboratively producing such situations is easy, but this strategy
  is not likely a stable equilibrium: Players can often gain a
  substantial advantage by going ``off script'' and instead trying to
  win the game. Additionally, sometimes the terms of
  chess-to-the-death do not allow the players to communicate at all
  beforehand, nor during the game. If this is the case, then it may be
  difficult to agree on the approach to drawing, let alone establish
  that this is both player's desire. Since the rules of
  chess-to-the-death can't forbid us from colluding at this moment, I
  hereby declare that the following is the correct approach: 
\begin{enumerate}[label=\arabic*.]
  \item[1.] Nf3. This is a reasonable opening move for white (begins the R\'eti)
    which can transpose into several common systems (e.g. King's
    Indian). Since the knight can move back to g1 on the next move,
    knight moves are the fastest route to a draw by repetition. This
    has a good chance of signaling to a wise player that a draw is
    desired. The player should make this move after pondering carefully
    for some time, and then looking meaningfully into the other player's
    eyes. 
  \item[1.] ... Nf6. This is both a strong response for black in a real game,
    and simultaneously a signal that a draw is desired. The other
    advantage is that very weak players may simply copy what white does.
    In doing so, they will also play this move. 
  \item[2.] Ng1?!. White moves the knight back to its starting square.
    This is a terrible move for white, but clearly signals the intention to
    draw.
  \item[2.] ... Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
    for a second time.
  \item[3.] Nf3 Nf6. At this point the game should clearly continue repeating
    the sequence, although since we are in the start position, white has
    any number of strong opening moves available. Signaling the draw line
    and then 3. d4!? may be pyschologically devastating.
  \item[4.] Ng1 Ng8 1/2-1/2. The starting position is reached for the third
    time, which by rule\footnotemark\ is a draw.
\end{enumerate}

@@

  If 1. ... Nc6 or another Knight's move, white can also consider
  continuing in the obvious way. However after 1. ... d5, black has
  refused or not noticed the draw. Fortunately, white is still in a
  good position to play the game normally (this is the main line of
  the R\'eti opening, followed by 2. c4). White can try to be more
  obvious with 2. Ng1, but if black is choosing to just play normally,
  white takes a distinct handicap by doing so.

  The biggest risk for white is that black does not play 2. ... Ng8
  but rather a normal move like 2. ... g6 (``Fool's Draw Betrayed'').
  This can happen if black is not metagaming at all (for Nf6 is a
  normal response to the normal Nf3), or if black is an exceptionally
  shrewd metagamer (tricking white into wasting two tempos with Ng1
  by pretending to be cooperating).

@@

For the sake of this question, we'll consider the the white king to
``die'' if white loses (i.e., is checkmated), and likewise for black.
Otherwise, of course, the best chances of survival would trivially be
with the two kings, since they always survive. Loss includes
resignation, since most high-level games actually end once the
defeated player agrees that loss is inevitable. We can think of this
common case like king seppuku. Many games also end in time forfeit,
which is like the king's poor diet and lifestyle choices leading to
a death by natural causes.

Neither side has a decisive advantage, and many games end in a draw,
with both kings surviving. So the survival chances of a king are
clearly greater than 50\%---pretty decent odds. Is it possible that
any other piece has even better chances?

\footnotetext{
  But is it? The lichess implementation (although known
  to be buggy\cite{kingme}) does not permit a threefold repetition
  claim in this situation, which got me thinking that maybe there is
  some subtlety here. Is the starting position special somehow, not
  counting as having occurred? The relevant statute, from the FIDE
  Laws of Chess\cite{fiderules}:
  \begin{quote}
    {\bf 9.2}.\quad The game is drawn, upon a correct claim by a
    player having the move, when the same position for at least the
    third time (not necessarily by a repetition of moves):
    \begin{enumerate}[label=\alph*.]
    \item is about to appear, if he first writes his move, which
      cannot be changed, on his scoresheet and declares to the arbiter
      his intention to make this move, or
    \item has just appeared, and the player claiming the draw has the
      move.
  \end{enumerate}

@@

Okay, this is pretty much off the cuff, but let’s say, just for white,
from most to least likely to survive:

@@

I figure the king has got to be somewhere near the middle of the pack
since he dies in half of games featuring a winner — but with slightly
higher-than-even odds of surviving, since some games end in a draw.
I’m probably mixing up means and medians here somehow..

@@

I’m gonna assume castling happens more often on the King’s side, so
let’s give Kingside Rook and F, G, and H Pawns a better shot than
their fellows on the left. But maybe it should actually be worse,
since if they die, it’s because they failed to protect the king. Plus,
having heard the tip about C Pawn loud and clear, I’m gonna assume
that bad boy most often becomes a new Queen, which means he gets more
survival points than the real Queen herself.

@@

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a
good example of a piece that has very different survival rates in
different opening preferences. Since the variance is otherwise very
low, I did not attempt to

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%.

@@

Other than the bishops---which have no legal way to reach half of the
squares---every piece reaches every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

@@

>>>>

If you are forced to play chess to the death, you are in trouble,
because most people are not good at chess (for example, the author)
and yet want to live.\footnote{It is easy for two players to
  collaborate to produce a draw, especially by simply agreeing to a
  draw at the outset of the game (if allowed). Some tournament formats
  forbid the players from agreeing to a draw verbally before a certain
  point (e.g. 30 moves), or without the arbiter's consent, and FIDE
  rules technically do not allow a draw until both players have made a
  move (5.2.3). There are always other routes to a draw, for example
  by stalemate or repeating the same position three times.
  Collaboratively producing such situations is easy, but this strategy
  is not likely a stable equilibrium: Players can often gain a
  substantial advantage by going ``off script'' and instead trying to
  win the game. Additionally, sometimes the terms of
  chess-to-the-death do not allow the players to communicate at all
  beforehand, nor during the game. If this is the case, then it may be
  difficult to agree on the approach to drawing, let alone establish
  that this is both players' desire. Since the rules of
  chess-to-the-death can't forbid us from colluding right now as you
  read this paper, I hereby declare that the following is the correct
  approach:
\begin{enumerate}[label=\arabic*.]
  \item[1.] Nf3. This is a reasonable opening move for white (begins the R\'eti)
    which can transpose into several common systems (e.g. King's
    Indian). Since the knight can move back to g1 on the next move,
    knight moves are the fastest route to a draw by repetition. This
    has a good chance of signaling to a wise player that a draw is
    desired. The player should make this move after pondering carefully
    for some time, and then looking meaningfully into the other player's
    eyes. 
  \item[1.] \ldots Nf6. This is both a strong response for black in a real game,
    and simultaneously a signal that a draw is desired. The other
    advantage is that very weak players may simply copy what white does.
    % XXX potential citation to 'elo world' if players are stateful..
    In doing so, they will also play this move. 
  \item[2.] Ng1?!. White moves the knight back to its starting square.
    This is a terrible move for white, but clearly signals the intention to
    draw.
  \item[2.] \ldots Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
    for a second time.
  \item[3.] Nf3 Nf6. At this point the game should clearly continue repeating
    the sequence, although since we are in the start position, white has
    any number of strong opening moves available. Signaling the draw line
    and then 3. d4!? may be pyschologically devastating.
  \item[4.] Ng1 Ng8 1/2-1/2. The starting position is reached for the third
    time, which by rule\footnotemark\ is a draw.
\end{enumerate}

@@

  If 1. \ldots Nc6 or another Knight's move, white can also consider
  continuing in the obvious way. However after 1. \ldots d5, black has
  refused or not noticed the draw. Fortunately, white is still in a
  good position to play the game normally (this is the main line of
  the R\'eti opening, followed by 2. c4). White can try to be more
  obvious with 2. Ng1, but if black is choosing to just play normally,
  white takes a distinct handicap by doing so.

  The biggest risk for white is that black does not play 2. \ldots Ng8
  but rather a normal move like 2. \ldots g6 (``Fool's Draw Betrayed'').
  This can happen if black is not metagaming at all (for Nf6 is a
  normal response to the normal Nf3), or if black is an exceptionally
  shrewd metagamer (tricking white into wasting two tempos with Ng1
  by pretending to be cooperating).

@@

For the sake of this question, we'll consider the the white king to
``die'' if white loses (i.e., is checkmated), and likewise for black.
Otherwise, of course, the best chances of survival would trivially be
with the two kings, since they are never formally captured. Loss
includes resignation, since most high-level games actually end once
the defeated player agrees that loss is inevitable. We can think of
this common case like king seppuku. Many games also end in time
forfeit, which is like the king's poor diet and lifestyle choices
leading to a death by natural causes.

Neither side is believed to have a decisive advantage, and many games
end in a draw, with both kings surviving. So the survival chances of a
king are likely greater than 50\%; pretty decent odds. Is it possible
that any other piece has even better chances? Let's find out---our
lives may depend on it!

\footnotetext{
  But is it?

  First of all, although either player is allowed to {\it claim} a
  draw after three repetitions of the same position, it is not
  automatic. However, FIDE rules do declare that the game simply ends
  in a draw upon {\it five} repetitions. Of course it is easy to
  extend the Fool's Draw to accommodate this.

  Second: The lichess implementation (although known to be
  buggy\cite{kingme}) does not permit a threefold repetition claim in
  this situation, which got me thinking that maybe there is some
  subtlety here. Is the starting position special somehow, not
  counting as having occurred? The relevant statute, from the FIDE
  Laws of Chess\cite{fiderules}:
  \begin{quote}
    {\bf 9.2}.\quad The game is drawn, upon a correct claim by a
    player having the move, when the same position for at least the
    third time (not necessarily by a repetition of moves):
    \begin{enumerate}[label=\alph*.]
    \item is about to appear, if he first writes his move, which
      cannot be changed, on his scoresheet and declares to the arbiter
      his intention to make this move, or
    \item has just appeared, and the player claiming the draw has the
      move.
  \end{enumerate}

@@

Okay, this is pretty much off the cuff, but let's say, just for white,
from most to least likely to survive:

@@

I figure the king has got to be somewhere near the middle of the pack
since he dies in half of games featuring a winner---but with slightly
higher-than-even odds of surviving, since some games end in a draw.
I’m probably mixing up means and medians here somehow..

@@

I’m gonna assume castling happens more often on the King’s side, so
let’s give Kingside Rook and F, G, and H Pawns a better shot than
their fellows on the left. But maybe it should actually be worse,
since if they die, it's because they failed to protect the king. Plus,
having heard the tip about C Pawn loud and clear, I’m gonna assume
that bad boy most often becomes a new Queen, which means he gets more
survival points than the real Queen herself.

@@

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
and declining the pawn are both popular and sound responses. This may
be a good example of a piece that has substantially different survival
rates in different opening preferences. Since the variance is
otherwise extremely low, I only report means for the remainder of the
paper.

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%. Although the database contains games in many time formats and
with all varieties of human skill (including over a thousand games by
Magnus Carlsen, the world champion and highest rated player of all
time\footnote{Although to be fair, his username ``DrDrunkenstein''
  suggests he may not play at full strength.}), blitz ($\sim 5$ minutes
per side) and bullet ($\sim 1$ minute) games are predominant. In
Section~\ref{sec:comparative} I show some slices of the data for
comparison. % XXX could summarize whether this is a real effect or
% not, once I have the slices!

@@

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

@@

\section{Guesses and slices}
%  
%  \begin{tabular}{|l|l|}
%  \hline
%    tom & actual \\
%   \hline
%   \Pawn f    &  \\
%   \Pawn c    &  \\
%   \Pawn g    &  \\
%   \Pawn a    &  \\
%   \Pawn h    &  \\
%   \Pawn b    &  \\
%   \Rook h    &  \\
%   \Rook a    &  \\
%   \King      &  \\
%   \Queen     &  \\
%   \Bishop f  &  \\
%   \Bishop c  &  \\
%   \Knight g  &  \\
%   \Knight b  &  \\
%   \Pawn e    &  \\
%   \Pawn d    &  \\
%  \hline
%  \end{tabular}
%
